\documentclass[12pt, twoside, a4paper]{report}

\author{Mayank Jatav}
\usepackage{mathptmx, lettrine}
\usepackage{mathtools,amsmath}
\usepackage{gensymb}
% Beautiful Math Font
\usepackage{eulervm}
\usepackage{mathptmx, lettrine}
\usepackage{mathtools,amsmath}
\usepackage{tabularray}
\usepackage{gensymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsfonts}

% Table Options
\usepackage{array, longtable, subcaption, multirow, tabularx, booktabs}
\captionsetup{%
	figurename=Figure
	tablename=Table
}
% Graphics and Figure Options
\usepackage{graphics, float, xcolor}

% Aligning/Formatting Equations
\usepackage{amsmath}

% Typeset Text
\usepackage{microtype}

% Customize Lists
\usepackage{enumitem}%\setlist[itemize]{noitemsep, topsep=0pt}
\setlist[itemize]{topsep=2pt}
\setlist[enumerate]{topsep=2pt}

\usepackage{lscape, rotating, layout}
\usepackage[a4paper, left=20mm, right=20mm, top=30mm, bottom=30mm]{geometry}	% Online
%\usepackage[a4paper, left=35mm, right=20mm, top=30mm, bottom=30mm]{geometry} 	% Print

% Fancy Header
\usepackage{fancyhdr}
\fancyhf{}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\setlength{\headheight}{3mm}
\setlength{\footskip}{10mm}

% Referencing Options
\usepackage{cite}
\usepackage{lscape}
% Blank Pages - Empty Header
\usepackage{emptypage}
% Chapter Title Design
\usepackage[T1]{fontenc}
\usepackage[charter]{quotchap}

% Hyphenation
\usepackage{hyphenat}
\usepackage[english]{babel}

% Color HyperLinks
\definecolor{darkpastelgreen}{rgb}{0.01, 0.75, 0.24}
\usepackage{hyperref, url}

% ...existing code...
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{0.5em}{0.5em}
\titlespacing*{\subsection}{0pt}{0.5em}{0.5em}
\titlespacing*{\subsubsection}{0pt}{0.5em}{0.5em}
% ...existing code...

% Online
\hypersetup{colorlinks=true, linkcolor=blue, citecolor=darkpastelgreen, filecolor=magenta, urlcolor=blue}
% Print
% \hypersetup{colorlinks=false, linkcolor=black, filecolor=black, urlcolor=black}
% Acronyms/Glossary
\usepackage[nopostdot, style=super, nonumberlist, toc, automake, acronym]{glossaries}
\makeglossaries

\renewcommand{\glsnamefont}[1]{\textbf{#1}}
\input{contents/glossary}

% Macro
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

% Big O Notation
\newcommand{\bigO}[1]{\ensuremath{\mathop{}\mathopen{}\mathcal{O}\mathopen{}\left(#1\right)}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\parskip 2.5mm		% Space before a new Paragraph
\parindent 0pt		% Indent Paragraphs
\linespread{1.3}	% Space between Lines of the paragraph

\setlength{\headheight}{15pt}

% Initialize Table of Contents, List of Figures and List of Tables
\AtBeginDocument{\addtocontents{toc}{\protect\thispagestyle{empty}}}
\AtBeginDocument{\addtocontents{lof}{\protect\thispagestyle{empty}}}
\AtBeginDocument{\addtocontents{lot}{\protect\thispagestyle{empty}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Change Page Numbering Format
\pagenumbering{arabic}
\pagestyle{fancy}

% Page Headings and Subheadings with Page Number
% TwoPage
\fancyhead[EL, OR]{\thepage}
\fancyhead[ER]{\nouppercase{\leftmark}}
\fancyhead[OL]{\nouppercase{\rightmark}}
% One Page
%\fancyhead[L, R]{\thepage}

\renewcommand{\sectionmark}[1]{\markright{\thesection\ #1}}
%\lhead{\fancyplain{}{\nouppercase{\rightmark}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\include{synopsis_contents/title_page}

\renewcommand{\thesection}{\arabic{section}}
\setcounter{secnumdepth}{10}

\section*{Introduction}

This synopsis presents a dual-track academic and practical contribution carried out as part of the Master’s thesis. The work is divided into two major components: \textbf{(i) Research Work} focusing on fabric classification using deep learning techniques, and \textbf{(ii) Internship Work} involving the design and development of a Security Orchestration, Automation, and Response (SOAR) platform during an industrial internship at Bharat Electronics Limited (BEL).

The research work addresses the challenge of accurate fabric identification, which is essential in textile manufacturing, quality assurance, and material categorization. A hybrid deep learning architecture combining Convolutional Neural Networks (CNNs) and Vision Transformers (ViTs) is proposed to enhance the classification of textile fabrics using both macro (RGB) and Optical Coherence Tomography (OCT) images. The integration of local and global feature representations enables robust and precise classification across multiple fabric types.

The internship work focuses on the implementation of a custom SOAR system designed to streamline security operations and automate incident response workflows in a Security Operations Center (SOC). The platform comprises key modules such as incident management, workflow orchestration, dashboard analytics, playbook automation, and MITRE ATT\&CK matrix integration. The system was developed to support real-time monitoring, consistent threat handling, and improved operational efficiency.

Together, these two efforts demonstrate the application of advanced engineering principles in solving real-world problems across manufacturing and cybersecurity domains, contributing both to academic research and industry-grade system development.

% \part{Fabric Classification using Deep Learning}

\input{synopsis_contents/research}
\input{synopsis_contents/internship}

% \part{Security Orchestration Automation and Response}

% References
\baselineskip=16pt	% Length between two lines in a paragraph

\addcontentsline{toc}{section}{References}
\bibliographystyle{unsrt}
\renewcommand{\bibname}{References}
\bibliography{synopsis_contents/references}

\end{document}