\chapter*{Acknowledgement}

Throughout the course of my Master's journey, I have been fortunate to receive guidance, encouragement, and support from several individuals, without whom this thesis would not have been possible. I take this opportunity to express my heartfelt gratitude to all those who have been an integral part of this academic endeavor.

First and foremost, I would like to express my deepest gratitude to my supervisor, \textbf{Prof. Pritee Khanna}, for her constant support, mentorship, and insightful guidance throughout the course of this work. Her expertise, constructive feedback, and patient encouragement have played a vital role in shaping the direction and quality of this research. I am also sincerely thankful to \textbf{Prof. Aparajita Ojha} for her consistent guidance and valuable suggestions. Her support, both academically and personally, helped me maintain a strong foundation throughout the research process.

I would like to extend my sincere thanks to my internship mentor, \textbf{Mr. Gaurav Damri}, for his mentorship, technical guidance, and for providing me the opportunity to work on real-world problems as part of my internship at Bharat Electronics Limited.

I am also thankful to my thesis committee members—\textbf{Dr. Avinash Chandra Pandey}, \textbf{Dr. Shivansh Mishra}, and \textbf{Dr. Nitish Andola}—for their time, insights, and constructive feedback which greatly helped in improving the quality of this work.

Beyond academic support, this journey was made easier and more fulfilling thanks to the unwavering encouragement and companionship of my peers. I would like to express heartfelt appreciation to \textit{Mohammad Arif Khan}, \textit{Amit Bhati}, \textit{Poornima Singh Thakur}, \textit{Shubhangi Chaturvedi}, \textit{Janvi Gohil}, \textit{Sudha Singh}, and \textit{Aman Yerwarkar} for their constant moral support, discussions, and uplifting presence.

Lastly, I owe my deepest gratitude to my family, for their unconditional love, sacrifices, and support throughout my life. Their faith in me has been my greatest strength.

\bigskip

\hfill\textit{--- Mayank Jatav}
